---
title: "Statistical Inference Project Part 1"
author: "Paulo Cardoso"
date: "January 28, 2016"
output: pdf_document
---

##Overview

This first part of Statistical Inference class project that aims to simulate an exponential distribution and perform comparisons with the central limit theorem and is composed of four stages:
  - Simulations
  - Sample Mean versus Theoretical Mean
  - Sample Variance versus Theoretical Variance
  - Distribution
The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of the exponential distribution is 1/lambda and the standard deviation is also 1/lambda. For this simulation, we set lambda=0.2. And we investigate the distribution of averages of 40 samples drawn from the exponential distribution with lambda=0.2.

## Simulations

Given the requirements of the exercises were held 1000 simulations using a distribution of averages of 40 and a lambda value of 0.2, as presented by de code below.
```{r echo=TRUE, cache=TRUE}
# Set seed
set.seed(456)
# Set lambda
lambda <- 0.2
# 1000 simulations with 40 samples 
sampleSize <- 40
nSim <- 1000

# Performing the 1000 simulations
simExp <- matrix(rexp(nSim*sampleSize, rate=lambda), nSim, sampleSize)
# Averages of 40 exponentials
rMean <- rowMeans(simExp)
``` 
The vector rMean contains the mean value of 40 samples.  And below you can observe its summary.
```{r echo=TRUE, cache=TRUE}
# Presenting Summary
summary(rMean)
``` 

## Sample Mean versus Theoretical Mean

Presentation of the distribtion of the sample mean and the theoretical mean.
```{r echo=TRUE, cache=TRUE}
# Sample mean
sampleMean <- mean(rMean)
sampleMean

# Theoretical mean
tMean <- 1/lambda
tMean

# Plot histogram of averages
hist(rMean, breaks=50, prob=TRUE,
     main="Distribution of averages of samples,
     drawn from exponential distribution with lambda=0.2",
     xlab="")
# Density of averages of the samples
lines(density(rMean))
# Theoretical center of distribution
abline(v=1/lambda, col="red")
# Theoretical density of the averages of samples
xfit <- seq(min(rMean), max(rMean), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sampleSize)))
lines(xfit, yfit, pch=22, col="green", lty=2)
# Adding legend
legend('topright', c("simulation", "theoretical", "theoretical mean"), lty=c(1,2,1), col=c("black", "green", "red"))
``` 

## Sample Variance versus Theoretical Variance

This is the comparison of theoretical variance with the sample variance and the theoretical standard error with the sample standard error.
```{r echo=TRUE, cache=TRUE}
# Theoretical variance
tVar <- (1/lambda)^2/sampleSize
round(tVar,3)

# Sample variance
round(var(rMean), 3)

# Theoretical standard error
tse <- 1/(lambda*sqrt(sampleSize))
round(tse, 3)

# Sample standard error
round(sd(rMean), 3)

# qqplot visualization
qqnorm(rMean); qqline(rMean)
``` 

## Distribution

The plot below shows that the distribution is approximately normal.
```{r echo=TRUE, cache=TRUE}
library(ggplot2)

vis <- data.frame(rMean)
a <- ggplot(vis, aes(x = rMean))
a <- a + geom_histogram(aes(y=..density..), colour="black")
a + geom_density(colour="red", size=1)
```
