---
title: "Statistical Inference Project Part 2"
author: "Paulo Cardoso"
date: "January 28, 2016"
output: pdf_document
---

# Overview

This is the second part of Statistical Inference class project that aims to perform a basic inferential analyses utilizing the ThoothGrowth data of the R datasets package.The dataset is compose by 60 observations, length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1 and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

# Set up

Loading the external libraries and the data.
```{r echo=TRUE, cache=TRUE}
# Loading libraries
library(datasets)
library(lattice)

# Loading the data
data(ToothGrowth)
``` 

# Exploratory Data Analyses

In order to understand the data set is necessary that a few metrics be presented.

I)Presenting the struture of the dataset.
```{r echo=TRUE, cache=TRUE}
# Struture presentation
str(ToothGrowth)
``` 
II)Display the amount of rows.
```{r echo=TRUE, cache=TRUE}
# Number of rows
nrow(ToothGrowth)
```
III)Display the first six rows.
```{r echo=TRUE, cache=TRUE}
# Head of the dataset
head(ToothGrowth)
```
IV)Display the last six rows.
```{r echo=TRUE, cache=TRUE}
# Tail of the dataset
tail(ToothGrowth)
```
V) Presenting the summary of the dataset.
```{r echo=TRUE, cache=TRUE}
# Summary
summary(ToothGrowth)
```
VI) Scatterplot showing every single value, and a boxplot chart summarizing the dataset.
```{r echo=TRUE, cache=TRUE}
# Scatterplot
xyplot(len~dose|supp, ToothGrowth,
       main="Scatterplots of type of supplement and dose",
       ylab="Length", xlab="Dose")
```
VII) Boxplot representing the tooth length by dose, on that way comparing the two delivery methods.And also a basic summary of each delivery method
```{r echo=TRUE, cache=TRUE}
# Separating the two diferent delivery methods
oj <- levels(ToothGrowth$supp)[1]
vc <- levels(ToothGrowth$supp)[2]

# Boxplot
par(mfrow = c(1, 2))
boxplot(len ~ dose, 
        ToothGrowth[ToothGrowth$supp == oj,], 
        main="OJ", 
        xlab="Dose", 
        ylab="Tooth Length", 
        ylim=range(0,35))
boxplot(len ~ dose, 
        ToothGrowth[ToothGrowth$supp == vc,], 
        main="VC", 
        xlab="Dose", 
        ylab="Tooth Length", 
        ylim=range(0,35))

# Summary from Orange Juice
summary(ToothGrowth[ToothGrowth$supp == oj,])

# Summary from Ascorbic Acid
summary(ToothGrowth[ToothGrowth$supp == vc,])
```

# Inferential Statistics

Do the tooth length of the guinea pigs depends on delivery methods?
To analyze tooth growth, we carry out two-sample (orange juice vs. ascorbic acid) T-tests at each dosage level.

The T-test at 0.5 mg yields the following:
```{r echo=TRUE, cache=TRUE}
t.test(len ~ supp, ToothGrowth[ToothGrowth$dose == .5, ])
```

The T-test at 1 mg yields the following:
```{r echo=TRUE, cache=TRUE}
t.test(len ~ supp, ToothGrowth[ToothGrowth$dose == 1, ])
```

The T-test at 2 mg yields the following:
```{r echo=TRUE, cache=TRUE}
t.test(len ~ supp, ToothGrowth[ToothGrowth$dose == 2, ])
```

# Conclusion

From the T test performed above it can be concluded that from statistical significance p-value that the dosages of 0.5 mg and 1 mg of orange juice is more effective. Thus promoting greater dental growth that ascorbic acid. Analyzing the p-value of 2 mg can not be concluded that orange juice promotes dental growth more effectively than ascorbic acid.